Predicting the effects of character displacement on food-web dynamics

### 2018-05-21 ECD_Supp_Mat.Rmd
final draft that I will send to Seth and Dolph for feedback

---
title: "Predicting the effects of character displacement on food-web dynamics"
author: "Matt Barbour"
output:
  pdf_document:
    df_print: paged
    fig_caption: true
    citation_package: natbib
    includes:  
      in_header: preamble-latex.tex
bibliography: references_ECD_supp_mat.bib
---

@@

where $r_i$ represents the intrinsic growth rate of resource $i$, $K_i$ represents the carrying capacity of resource $i$, $e_{ji}$ represents the conversion efficiency of resource $i$ into new consumer $j$, and $m_j$ represents the mortality rate of consumer $j$. $F_{ji}(R_i)$ represents consumer $j$'s feeding rate (i.e. functional response) on resource $i$. This model is a useful characterization of a scenario where consumers compete for two distinct resources (e.g. zooplankton and benthic invertebrates) rather than if resources are better characterized by a continuous trait distribution (e.g. seed size, see @Taper1985 for an example). In the scenario where consumer feeding rates increase linearly with resource density, such that $F_{ji}(R_i)=a_{ji}R_i$ where $a_{ji}$ is the attack rate of consumer $j$ on resource $i$, then my model becomes MacArthur's model of resource competition (@MacArthur1972). The evolution of consumer attack rates in this model (and several extensions) have been analyzed in detail (@Lawlor1976; @Abrams1986). The general result of these analyses is that divergent character displacement in attack rates is an inevitable outcome of resource competition. <!-- In other words, the presence of a competitor causes consumers to evolve more specialized attack rates ($\frac{a_{ii}}{a_{ii}+a_{ij}}$) compared to when no competitor is present. -->

Here, I examine the effects of this inevitable divergence in attack rates on food-web dynamics. In particular, I examine how divergence in attack rates affects resource abundances as well as the local stability of both resources and consumers to perturbations. I do this by (i) deriving analytical expressions for the relationship between attack rates and food-web dynamics in different foraging scenarios and (ii) simulating the effects of competition on the eco-evolutionary dynamics of consumers and resources. For these simulations, I used an Adaptive Dynamics approach. At each evolutionary time step, I created a mutant consumer by randomly choosing a consumer and modifying its attack rate by either subtracting or adding a small constant <!-- (0.001 in the following simulations) -->with equal probability. Assuming this mutant consumer is rare, I then determined whether the mutant had higher relative fitness than the resident consumer, and thus could invade and replace the resident consumer. If the mutant was able to invade, I updated the attack rate of the resident consumer to the mutant attack rate and allowed consumer and resource dynamics to reach a steady state. I then repeated the simulation 10,000 times, which was sufficient for consumers to either reach an evolutionary stable strategy (ESS, @Smith1973) or an evolutionary limit (e.g. $\frac{a_{ii}}{a_{ii}+a_{ij}}$ is constrained to a maximum of 1 and minimum of 0). I show that the consequences of character displacement on food-web dynamics are not intuitive and depend critically on the foraging scenario and tradeoff in consumer attack rates.

## Effect of character displacement on resource abundances

I start by examing the effects of character displacement on the abundance of resources at equilibrium, which will also set the stage for a later analysis of food-web stability. In particular, I examine two different foraging scenarios: one where consumers forage for both resources simultaneously (e.g. @MacArthur1972) and one where they cannot because resources occur in distinct habitats (e.g. @Lawlor1976; @McCann2005). I say consumers have undergone divergent character displacement if there evolved attack rates are more specialized, defined as $|\frac{a_{ii}}{a_{ii}+a_{ij}}-0.5|$, when evolving with a competing consumer.

@@

An inherent assumption in MacArthur's model of resource competition is that consumers can forage for both resources simultaneously. In essence, this implies that both resources occur within the same habitat. To gain analytical insight to this scenario, I assume that resources are equivalent ($r=r_1=r_2$ and $K=K_1=K_2$) as well as consumers ($e=e_{11}=e_{12}=e_{21}=e_{22}$; $m=m_1=m_2$), except that consumer attack rates are mirror images of each other ($a_{ii}=a_{11}=a_{22}$; $a_{ij}=a_{12}=a_{21}$). While this scenario is arguably simplistic, it isolates the effect of character displacement rather than any dependence on inherent differences in consumers or resources. I show that when both consumers and resources are present that the abundance of resources at equilibrium are equivalent ($R=R_1=R_2$) and are determined by the following equation (*Mathematica* file with derivation available upon request):

$$\hat{R}=\frac{1}{a_{ii}+a_{ij}}\cdot\frac{m}{e}$$
Note that a key determinant of resource abundance is the consumer's total attack rate, $a_{ii}+a_{ij}$. Since the total attack rate is in the denominator, this implies that resource abundances will decrease if evolution favors an increase in total attack rate. I assume that consumers are not 'Darwinian demons' (@Law1979) and thus the evolution of their attack rates are subject to certain constraints. I modelled these constraints such that $\big(a_{ii}/{A}\big)^n+\big(a_{ij}/{A}\big)^n=1$, where $A$ is the total investment in attack rates and $n$ describes the shape of the tradeoff (@Sargent2006). This function has the useful property that it differentiates between cases where intermediate combinations of $a_{ii}$ and $a_{ij}$ are higher, on average, than the extremes (when $n>1$, green line in Fig. \ref{fig:Tradeoff}) or, conversely, where the two extremes are higher, on average, than intermediate investments (when $n<1$, orange line Fig. \ref{fig:Tradeoff}). When $n=1$, the tradeoff function is linear, and all combinations of $a_ii$ and $a_ij$ have the same total attack rate (blue line in Fig. \ref{fig:Tradeoff}).

@@

Previous work has shown that consumers still undergo divergent character displacement regardless of the shape of the tradeoff (@Lawlor1976; @Abrams1986). Interestingly, I find that the shape of the tradeoff qualitatively affects the relationship between character displacement and resource abundance (Fig. \ref{fig:MacArthur_Resources}). For example, if consumer's are constrained by a linear tradeoff ($n=1$), then there is no net change in total attack rate (blue points in Fig. \ref{fig:MacArthur_Resources}A) and character displacement has no effect on resource abundances (Fig. \ref{fig:MacArthur_Resources}B). If the tradeoff is concave down ($n>1$), then resource abundances can actually increase under character displacement (green points in Fig. \ref{fig:MacArthur_Resources}B). This is because the total attack rate of consumers is maximized at intermediate values ($a_{ii}=a_{aij}$) and decreases as consumers diverge (Fig. \ref{fig:MacArthur_Resources}A). Note that detecting this effect in nature would likely be difficult, since a concave-down tradeoff results in relatively small displacement relative to other tradeoff shapes (@Lawlor1976; green points in Fig. \ref{fig:MacArthur_Resources}A,B). When the tradeoff is concave up ($n<1$), character displacement suppresses resource densities due to the increase in total attack rates (orange points in Fig. \ref{fig:MacArthur_Resources}A,B). Although the equation I derived for resource abundance was for the scenario where both consumers and both resources are present, it predicts well the abundance of resources when a single consumer reaches its ESS (triangles on respective colored lines in Fig. \ref{fig:MacArthur_Resources}B). This is because a single consumer evolves to be a generalist that has equal attack rates on each resource (triangles at 0.5 along x-axis in Fig. \ref{fig:MacArthur_Resources}A), resulting in equivalent resource abundances.

@@

where $w_{ji}$ represents the proportion of time consumer $j$ spends foraging in a habitat where only resource $j$ is found (i.e. habitat preference). Note that since $w_ji$ is a proportion that $w_{jj}=1-w_{ji}$. As with attack rates, I assume that consumer habitat preferences are mirror images of each other ($w=w_{11}=w_{22}$).

@Lawlor1976 found that consumers still underwent divergent character displacement regardless of whether consumers could or could not forage for resources simultaneously. I find that the foraging scenario qualitatively affects the relationship between character displacement and resource abundances. To demonstrate this, note that again resource abundances are equivalent at equilibrium ($R=R_1=R_2$ when both consumers and resources present), but are now determined by the following equation (*Mathematica* file with derivation available upon request):

@@

This equation implies that if consumers evolve to become specialists on resources that occur in their preferred habitat (e.g. $w>0.5$ and $a_{ii}>a_{ij}$), then character displacement will always increase the effective attack rate of consumers ($wa_{ii}+(1-w)a_{ij}$), regardless of the tradeoff (Fig. \ref{fig:LS_Resources}A). Thus, character displacement always results in resource suppression if consumers are competing for resources that occur in different habitats (Fig. \ref{fig:LS_Resources}B). Note that the shape of the tradeoff can modify the effect of character displacement. This is not so much due to the tradeoff affecting the magnitude of displacement (it does, but the effect is minor), but because the form of the tradeoff affects resource abundances when a single consumer has reached its ESS (triangles in Fig. \ref{fig:LS_Resources}B). In contrast, resource abundances reach a similar value when consumers evolve in the presence of a competitor (circles in Fig. \ref{fig:LS_Resources}B), because character displacement tends to reach a constraint of complete specialization.

@@

It is worth noting that in this foraging scenario, resource abundances are consistently higher at the single consumer ESS compared to the predictions I derived for when both consumers are present (deviation of triangles from respective colored lines in Fig. \ref{fig:LS_Resources}B). This is likely because consumers actually evolve to be slightly specialized on the resources that occur in their non-preferred habitat (deviation of triangles from 0.5 along x-axis in Fig. \ref{fig:LS_Resources}B). <!-- I will revisit this subtle point when I examine the effects of character displacement on food-web stability. -->

@@

The effect of the foraging scenario on the relationship between character displacement and resource abundances appears to be quite general. For example, even if I consider a more realistic consumer functional response (derived by @McCann2005):

@@

I still observe the same qualitative relationship between character displacement and resource suppression (Fig. \ref{fig:McCann_Resources} in Appendix). This is because resource competition always leads to character divergence and equilibrium resource abundances are governed by similar dynamics  (*Mathematica* file with derivations available upon request):

@@

@Lawlor1976's work gave fundamental insight to the effect of character displacement on ecological stability; however, none of their stability analyses compared food webs with and without a competing consumer. Such comparisons are necessary for inferring the effects of ecological character displacement, a point that has been made clear in the criteria to demonstrate character displacement (@Schluter2000; @Schluter1992). In fact, the first two metrics of stability can only be assessed when both consumers are present because it examine the effect of divergence on coexistence, which requires both consumers to be present. In contrast, local stability can be compared in food webs with and without a competing consumer, and thus provides a useful benchmark for examining the effects of character displacement on food-web stability.

Below, I plot the results from the same simulations I did previously except now local stability ($-1\times\operatorname{Re}(\lambda_{max})$) is on the y-axis (Fig. \ref{fig:Stability}). In general, I find that resource supression goes hand-in-hand with local stability, a tendency that has been noted by others (@Murdoch2003; @McCann2011). Whenever character displacement results in resource suppression, I observe a decrease in local stability (Fig. \ref{fig:Stability}). This is not simply a consequence of having an additional consumer in the system, but emerges from the eco-evolutionary feedback between character displacement and resource suppression (compare initial (small points) with evolved strategies (large points), Fig. \ref{fig:Stability}). Note that I reach the same fundamental conclusions when I incorporated a more realistic functional response for the consumers (Fig. \ref{fig:Stability_McCann}).

@@

I condcuted the previous simulations on a scenario where the food web remains locally stable ($\operatorname{Re}(\lambda_{max})<0$). Is it possible that character displacement would ever result in a locally unstable food web? To gain analytical insight to this question, I examined the conditions governing the local stability of the four-species food web when consumers exhibit a more realistic functional response (@McCann2005). I show that the four-species food web will transition from having a locally stable equilibrium to a limit cycle under the following conditions (*Mathematica* file with derivation available upon request):

@@

Consumer and resource densities in these dynamical model can be interpreted from either a population or biomass perspective (@Yodzis1992; @Murdoch2003; @McCann2011). For the stickleback system, you have data on seasonal dynamics of resources in terms of both abundances and biomass. Within a season, I do not expect stickleback abundances to be coupled to the abundances of zooplankton and benthic invertebrates, given that stickleback reproduce annually whereas zooplankton and benthic invertebrates can reproduce many times within a season. Instead, I expect that stickleback biomass will be coupled to the biomass of zooplankton and benthic invertebrates within a season. This is because there will be lots of young stickleback at the beginning of the season, and they will eat a lot of resources and grow in size as the season progresses. If this is true, I also expect to see the largest negative effect of stickleback on resource densities late in the season.

>>>>

---
title: "Predicting the effects of character displacement on food-web dynamics"
author: "Matt Barbour"
output:
  pdf_document:
    df_print: paged
    fig_caption: true
    citation_package: natbib
    includes:  
      in_header: preamble-latex.tex
bibliography: references_ECD_supp_mat.bib
fontsize: 12pt
---

@@

where $r_i$ represents the intrinsic growth rate of resource $i$, $K_i$ represents the carrying capacity of resource $i$, $e_{ji}$ represents the conversion efficiency of resource $i$ into consumer $j$, and $m_j$ represents the mortality rate of consumer $j$. $F_{ji}(R_i)$ represents consumer $j$'s feeding rate (i.e. functional response) on resource $i$. This model is a useful characterization of a scenario where consumers compete for two distinct resources (e.g. zooplankton and benthic invertebrates) rather than if resources are better characterized by a continuous trait distribution (e.g. seed size, see @Taper1985 for an example). In the scenario where consumer feeding rates increase linearly with resource density, such that $F_{ji}(R_i)=a_{ji}R_i$ where $a_{ji}$ is the attack rate of consumer $j$ on resource $i$, then this model becomes MacArthur's model of resource competition (@MacArthur1972). The evolution of consumer attack rates in this model (and several extensions) have been analyzed in detail (@Lawlor1976; @Abrams1986). The general result of these analyses is that divergent character displacement in attack rates is an inevitable outcome of resource competition. <!-- In other words, the presence of a competitor causes consumers to evolve more specialized attack rates ($\frac{a_{ii}}{a_{ii}+a_{ij}}$) compared to when no competitor is present. -->

Here, I examine the effects of this inevitable divergence in attack rates on food-web dynamics. In particular, I examine how divergence in attack rates affect resource densities as well as the local stability of both resources and consumers. I do this by (i) deriving analytical expressions for the relationship between attack rates and food-web dynamics in different foraging scenarios and (ii) simulating the effects of competition on the eco-evolutionary dynamics of consumers and resources. For these simulations, I used an Adaptive Dynamics approach. At each evolutionary time step, I created a mutant consumer by randomly choosing a consumer and modifying its attack rate on one resource by either subtracting or adding a small constant <!-- (0.001 in the following simulations) -->with equal probability. The mutant's attack rate on the other resource was determined by the shape of the tradeoff (details below). Assuming this mutant consumer is rare, I then determined whether the mutant had higher relative fitness than the resident consumer, and thus could invade and replace the resident consumer. If the mutant was able to invade, I updated the attack rate of the resident consumer to the mutant attack rate and allowed consumer and resource dynamics to reach a steady state. I then repeated the simulation up to 10,000 times, which was sufficient for consumers to either reach an evolutionary stable strategy (ESS, @Smith1973) or an evolutionary limit (e.g. $\frac{a_{ii}}{a_{ii}+a_{ij}}$ is constrained to a maximum of 1 and minimum of 0). I show that the consequences of character displacement on food-web dynamics are not intuitive and depend critically on the foraging scenario and tradeoff in consumer attack rates.

## Effect of character displacement on resource densities

I start by examing the effects of character displacement on the density of resources at equilibrium, which will also set the stage for a later analysis of food-web stability. In particular, I examine two different foraging scenarios: one where consumers forage for both resources simultaneously (e.g. @MacArthur1972) and one where they cannot because resources occur in distinct habitats (e.g. @Lawlor1976; @McCann2005). I say consumers have undergone divergent character displacement if there evolved attack rates are more specialized, defined as $|\frac{a_{ii}}{a_{ii}+a_{ij}}-0.5|$, when evolving with a competing consumer.

@@

An inherent assumption in MacArthur's model of resource competition is that consumers can forage for both resources simultaneously. In essence, this implies that both resources occur within the same habitat. To gain analytical insight to this scenario, I assume that resources are equivalent ($r=r_1=r_2$ and $K=K_1=K_2$) as well as consumers ($e=e_{11}=e_{12}=e_{21}=e_{22}$; $m=m_1=m_2$), except that consumer attack rates are mirror images of each other ($a_{ii}=a_{11}=a_{22}$; $a_{ij}=a_{12}=a_{21}$). While this scenario is arguably simplistic, it isolates the effect of character displacement rather than any dependence on inherent differences in consumers or resources. I show that when both consumers and resources are present that the density of resources at equilibrium are equivalent ($R=R_1=R_2$) and are determined by the following equation (*Mathematica* file with derivation available upon request):

$$\hat{R}=\frac{1}{a_{ii}+a_{ij}}\cdot\frac{m}{e}$$
Note that a key determinant of resource density is the consumer's total attack rate, $a_{ii}+a_{ij}$. Since the total attack rate is in the denominator, this implies that resource densities will decrease if evolution favors an increase in total attack rate. I assume that consumers are not 'Darwinian demons' (@Law1979) and thus the evolution of their attack rates are subject to certain constraints. I modelled these constraints such that $\big(a_{ii}/{A}\big)^n+\big(a_{ij}/{A}\big)^n=1$, where $A$ is the total investment in attack rates and $n$ describes the shape of the tradeoff (@Sargent2006). This function has the useful property that it differentiates between cases where intermediate combinations of $a_{ii}$ and $a_{ij}$ are higher, on average, than the extremes (when $n>1$, green line in Fig. \ref{fig:Tradeoff}) or, conversely, where the two extremes are higher, on average, than intermediate investments (when $n<1$, orange line Fig. \ref{fig:Tradeoff}). When $n=1$, the tradeoff function is linear, and all combinations of $a_{ii}$ and $a_{ij}$ have the same total attack rate (blue line in Fig. \ref{fig:Tradeoff}).

@@

Previous work has shown that consumers undergo divergent character displacement regardless of the shape of the tradeoff (@Lawlor1976; @Abrams1986). Interestingly, I find that the shape of the tradeoff qualitatively affects the relationship between character displacement and resource density (Fig. \ref{fig:MacArthur_Resources}). For example, if consumer's are constrained by a linear tradeoff (blue lines), then there is no net change in total attack rate (Fig. \ref{fig:MacArthur_Resources}A) and character displacement has no effect on resource densities (Fig. \ref{fig:MacArthur_Resources}B). If the tradeoff is concave down (green lines), then resource abundances can actually increase under character displacement (Fig. \ref{fig:MacArthur_Resources}B). This is because the total attack rate of consumers is maximized at intermediate values ($a_{ii}=a_{ij}$) and decreases as consumers diverge (Fig. \ref{fig:MacArthur_Resources}A). Note that detecting this effect in nature would likely be difficult, since a concave-down tradeoff results in relatively small displacement relative to other tradeoff shapes (@Lawlor1976; green points in Fig. \ref{fig:MacArthur_Resources}A,B). When the tradeoff is concave up (orange lines), character displacement suppresses resource densities due to the increase in total attack rates (Fig. \ref{fig:MacArthur_Resources}A,B). Although the equation I derived for resource density was for the scenario where both consumers and both resources are present, it predicts well the density of resources when a single consumer reaches its ESS (triangles on respective colored lines in Fig. \ref{fig:MacArthur_Resources}B). This is because a single consumer evolves to be a generalist that has equal attack rates on each resource (triangles at 0.5 along x-axis in Fig. \ref{fig:MacArthur_Resources}A), resulting in equivalent resource densities.

@@

where $w_{ji}$ represents the proportion of time consumer $j$ spends foraging in a habitat where only resource $j$ is found (i.e. habitat preference). Note that since $w_{ji}$ is a proportion that $w_{jj}=1-w_{ji}$. As with attack rates, I assume that consumer habitat preferences are mirror images of each other ($w=w_{11}=w_{22}$).

@Lawlor1976 found that consumers still underwent divergent character displacement regardless of whether consumers could or could not forage for resources simultaneously. I find that the foraging scenario qualitatively affects the relationship between character displacement and resource densities. To demonstrate this, note that again resource densities are equivalent at equilibrium ($R=R_1=R_2$ when both consumers and resources present), but are now determined by the following equation (*Mathematica* file with derivation available upon request):

@@

This equation implies that if consumers evolve to become specialists on resources that occur in their preferred habitat (e.g. $w>0.5$ and $a_{ii}>a_{ij}$), then character displacement will always increase the effective attack rate of consumers ($wa_{ii}+(1-w)a_{ij}$), regardless of the tradeoff (Fig. \ref{fig:LS_Resources}A). Thus, character displacement always results in resource suppression if consumers are competing for resources that occur in different habitats (Fig. \ref{fig:LS_Resources}B). Note that the shape of the tradeoff can modify the effect of character displacement. This is not so much due to the tradeoff affecting the magnitude of displacement (it does, but the effect is minor), but because the form of the tradeoff affects resource abundances when a single consumer has reached its ESS (triangles in Fig. \ref{fig:LS_Resources}B). In contrast, resource densities reach a similar value when consumers evolve in the presence of a competitor (circles in Fig. \ref{fig:LS_Resources}B), because character displacement tends to reach a constraint of complete specialization.

@@

It is worth noting that in this foraging scenario, resource densities are consistently higher at the single consumer ESS compared to the predictions I derived for when both consumers are present (deviation of triangles from respective colored lines in Fig. \ref{fig:LS_Resources}B). This is likely because consumers actually evolve to be slightly specialized on the resources that occur in their non-preferred habitat (deviation of triangles from 0.5 along x-axis in Fig. \ref{fig:LS_Resources}B). <!-- I will revisit this subtle point when I examine the effects of character displacement on food-web stability. -->

@@

The effect of the foraging scenario on the relationship between character displacement and resource densities appears to be quite general. For example, even if I consider a more realistic consumer functional response (derived by @McCann2005):

@@

I still observe the same qualitative relationship between character displacement and resource suppression (Fig. \ref{fig:McCann_Resources} in Appendix). This is because resource competition always leads to character divergence and equilibrium resource densities are governed by similar dynamics  (*Mathematica* file with derivations available upon request):

@@

@Lawlor1976's work gave fundamental insight to the effect of character displacement on ecological stability; however, none of their stability analyses compared food webs with and without a competing consumer. Such comparisons are necessary for inferring the effects of ecological character displacement, a point that has been made clear in the criteria to demonstrate character displacement (@Schluter2000; @Schluter1992). In fact, the first two metrics of stability can only be assessed when both consumers are present because it examines the effect of divergence on coexistence, which requires both consumers to be present. In contrast, local stability can be compared in food webs with and without a competing consumer, and thus provides a useful benchmark for examining the effects of character displacement on food-web stability.

Below, I plot the results from the same simulations I did previously except now local stability ($-1\times\operatorname{Re}(\lambda_{max})$) is on the y-axis (Fig. \ref{fig:Stability}). In general, I find that resource supression goes hand-in-hand with local stability, a tendency that has been noted by others (@Murdoch2003; @McCann2011). Whenever character displacement results in resource suppression, I observe a decrease in local stability (Fig. \ref{fig:Stability}). This is not simply a consequence of having an additional consumer in the system, but emerges from the eco-evolutionary feedback between character displacement and resource suppression (compare initial (small points) with evolved strategies (large points), Fig. \ref{fig:Stability}). Note that I reach the same fundamental conclusions when I incorporated a more realistic functional response for the consumers (Fig. \ref{fig:Stability_McCann} in Appendix).

@@

I conducted the previous simulations on a scenario where the food web remains locally stable ($\operatorname{Re}(\lambda_{max})<0$). Is it possible that character displacement would ever result in a locally unstable food web? To gain analytical insight to this question, I examined the conditions governing the local stability of the four-species food web when consumers exhibit a more realistic functional response (@McCann2005). I show that the four-species food web will transition from having a locally stable equilibrium to a limit cycle under the following conditions (*Mathematica* file with derivation available upon request):

@@

Consumer and resource densities in these dynamical models can be interpreted from either a population or biomass perspective (@Yodzis1992; @Murdoch2003; @McCann2011). For the stickleback system, you have data on seasonal dynamics of resources in terms of both abundances and biomass. Within a season, I do not expect stickleback abundances to be coupled to the abundances of zooplankton and benthic invertebrates, given that stickleback reproduce annually whereas zooplankton and benthic invertebrates can reproduce many times within a season. Instead, I expect that stickleback biomass will be coupled to the biomass of zooplankton and benthic invertebrates within a season. This is because there will be lots of young stickleback at the beginning of the season, and they will eat a lot of resources and grow in size as the season progresses. If this is true, I also expect to see the largest negative effect of stickleback on resource densities late in the season.
